Detection and partial characterisation of stellar multiplicity with Gaia

### 2018-08-29 article/ms.tex
some words

There are many ways that binaries can be reliably identified using \Gaia\
data, including radial velocity measurements, astrometry, and photometry.
These data are sensitive to characterising stellar multiplicity in 
different ways. Here we describe the data that we use from the second
\Gaia\ data release \citep{Lindegren:2018} before defining our model.

@@

For a single star, or a star without a significant mass companion, the source radial
velocity jitter represents the minimum uncertainty with which \Gaia\ can measure 
radial velocity for a star of that colour, apparent magnitude, and absolute magnitude.
Stars with bluer colours have fewer absorption lines with deeper wings, and fainter stars
have on average lower signal-to-noise (S/N) ratios, both of which result in noisier radial
velocity measurements. The absolute magnitude is similarly important, as giant stars have
narrow absorption lines than main-sequence stars of the same temperature and colour. For stellar multiples, the source radial velocity jitter will be the quadrature sum of the expected jitter for the primary (if it were a single star) and a contribution from the radial velocity semi-amplitude of the orbit.

Some sources in \Gaia\ do not have reported radial velocities even though they are
apparently bright enough and within a suitable \bp\ - \rp\ colour range.  The likely
explanation for these situations is that the \Gaia\  team identified
the source to be a double-lined spectroscopic binary, and there is no utility in
reporting a point estimate of the \emph{source} radial velocity and associated error.
Other explanations are also possible, which implies that we cannot use the \emph{lack}
of radial velocity as a robust identifier of stellar multiplicity. \todo{In the following
Sections}
% Show it ehre?
we demonstrate that for sources that do not have reported radial velocities -- and 
sources of similar apparent magnitude and colour \emph{do} have radial velocities --
their astrometric jitter is usually sufficient to distinguish single stars from stellar
multiples.

@@

Many higher-order star systems are unresolved point sources in \Gaia, but the
stellar companion(s) may have sufficient mass to measurably affect the tangential 
velocity of the primary and produce a detectable jitter in the astrometric 
position. 
%There are many other sources of astrometric jitter, but here we show 
%that the principle effect of astrometric excess noise is due to stellar 
%companions.
We make use of the astrometric unit weight error $u$ to measure astrometric
jitter
\begin{eqnarray}
	u = \sqrt{\frac{\chi^2_{al}}{N_{obs,al} - 5}} \label{eq:astrometric_unit_weight_error}
\end{eqnarray}
\noindent{}where $\chi^2_{al}$ is the astrometric $\chi^2$ value
in the along-scan direction (column \texttt{astrometric\_chi2\_al}) and 
$N_{obs,al}$ is the number of astrometric observations in the along-scan
direction (column \texttt{astrometric\_n\_obs\_al}). 
The astrometric unit weight error is unaffected by the `degree of freedom bug'
that occurred during the processing of the the second \Gaia\ data release
\citep[e.g., see Appendix A1 of ][]{Lindegren:2018}, which resulted in 80\% of
sources having zero astrometric excess noise. 
The astrometric unit weight error can be interpreted similarly to how the
astrometric excess noise would be interpreted (e.g., large values are less
consistent with a single star model), but using the astrometric unit weight
error allows us to distinguish astrometric binaries in a much larger sample.
And like the radial velocity jitter, the unit weight astrometric error of a
single source must be considered in context with stars of similar colour 
and apparent magnitude (but not necessarily absolute magnitude).

@@

We restrict our work to the brightest $\sim{}10^7$ sources in \Gaia\ DR2.
About 70\% of sources in this subset have reported radial velocities,
allowing us to identify spectroscopic binaries, and this subset provides
a good sample of sources to test the precision to which we can identify
astrometric binaries. Specifically, we executed the following Astronomical Data Query
Language \citep[\texttt{ADQL};][]{someone} query on \ESA\ \Gaia\ archive
\citep{esa_gaia_archive}:

@@

We assume that for a sample of sources with similar \bp\ - \rp\ colours
and apparent \rp\ magnitudes -- and absolute \rp\ magnitudes, if radial velocities
are being considered -- the distribution of jitter is a mixture of two components:

\begin{enumerate}
\item \emph{single-star systems}, where the jitter represents the minimum noise that 
	   \Gaia\ can measure for a source with those properties (e.g., a source with
	   that \bp\ - \rp\ colour, apparent \rp\ magnitude, and perhaps absolute \rp\
	   magnitude);

@@

We further assume that the jitter for a single star will change depending on the
\bp\ - \rp\ colour, the apparent magnitude, and the absolute magnitude. For example, 
the mean radial velocity jitter of single stars with $\teff \approx 6000\,\textrm{K}$
will be higher than the jitter of single stars with $\teff \approx 5000\,\textrm{K}$, 
with all else being (nearly) equal. Similarly, the minimum radial velocity jitter
for giant stars is expected to be lower on average than for main-sequence stars of
the same colour, as giant stars have narrower absorption lines than main-sequence
stars of the same colour, allowing for a more precise measurement of the radial
velocity. However, this effect will not affect the astrometric jitter: the minimum
astrometric jitter for a main-sequence star and a giant star of the same temperature
is likely to be indistinguishable. For these reasons, in order to evaluate whether 
a source is more likely to be a single star or a stellar multiple, we must consider 
the jitter in context with other sources that have a similar \bp\ - \rp\ colour, 
apparent \rp\ magnitude, and in the case of radial velocity jitter, the absolute
\rp\ magnitude.\footnote{The magnitude in any band would be sufficient. Throughout
this work we chose the \rp\ band because its wavelength range most closely overlaps 
with that of the Radial Velocity Spectrograph on board \Gaia.} However, we cannot reliably describe a quantitative
relationship for \emph{how} exactly the astrometric and radial velocity jitter is
expected to vary for a single star given some source properties. For these reasons
we chose to adopt a flexible non-parametric model for stellar multiplicity, primarily
conditioned on the assumption that if we select enough sources with \emph{similar}
properties then within that subset of similar stars there exists \emph{some} single stars 
and \emph{some} stellar multiples.

We illustrate our approach in Figure \ref{fig:npm_schematic}, which can be qualitatively
described by the following steps. When considering the radial velocity jitter for a
\Gaia\ source of interest, we select $N$ sources that have a similar \bp\ - \rp\ colour,
apparent \rp\ magnitude, and absolute \rp\ magnitude. We then fit the distribution of
radial velocity jitter $\RVJitter$ as a mixture of two components: a normal distribution
to represent single stars, and a log-normal distribution to represent stellar multiples.
We perform these steps for $\approx10^4$ sources. While this process approximately maps
how the radial velocity jitter of single (and multiple) stars varies across the Hertszprung-Russell
diagram, each source of interest is considered independent from all others. For example,
for three sources that are all indistinguishable from each other in \bp\ - \rp\ colour, 
apparent \rp\ magnitude, and absolute \rp\ magnitude, the procedure we described above
treats the fitting of these three mixture models as entirely independent of each other,
even though we suspect that the parameters of the three mixture models ought to be 
identical. In other words: the procedure as described does not enforce any smoothness.
Given enough data this is not a problem for densely-populated regions of the 
Hertszprung-Russell diagram as the optimized results \emph{will} be (statistically)
indistinguishable for similar sources. In practice this poses a problem for poorly-sampled
regions of the Hertzsprung-Russell diagram, and raises subjective questions on what
constitutes \emph{the most} similar type of source.

\todo{In the third step we apply a Gaussian process.}

@@

\subsection*{Step 1: Selecting similar sources}

We construct a $k$-dimensional tree \citep[k-d tree;][]{someone} for all sources in the
dimensions of \bp\ - \rp\ colour, apparent \rp\ magnitude, and absolute \rp\ magnitude.
When doing so we scale the dimensions such that \todo{0.1}\,mags in \bp\ - \rp\ colour is
approximately equal to \todo{1} magnitude in apparent \rp\ magnitude and \todo{1} magnitude in 
absolute \rp\ magnitude. We use this $k$-d tree for selecting a `ball' of similar
stars for a given source.

We enforce a number of constraints when selecting similar stars. For each source
we require the multi-dimensional `ball' to include at least \todo{128} sources.
We further require that the radius of the ball extends to at least \todo{0.05}\,mags in 
\bp\ - \rp\ colour, \todo{0.5}\,mags in apparent \rp\ magnitude, and \todo{0.5}\,mags in absolute
\rp\ magnitude. If the initial query of \todo{128} points does not extend wide enough to
meet our constraints on minimum radius, we extend the radius (linearly in each scaled)
dimension until the minimum constraints on radii are met. Our minimum radii constraints
imply that there will be many (e.g., $>>\todo{128}$) points returned in dense regions of the
Hertszprung-Russell diagram. In these situations we then subsample the selected points
to return a random \todo{1024} sources in any ball, while enforcing our constraints on
minimum radii. As a result, the number of `similar' sources selected will range from
a minimum of \todo{128} to a maximum of \todo{1024}, and the minimum radii constraints on various
dimensions are always maintained.

@@

\subsection*{Step 2: Fitting the two-component mixture}

We fit a two-component mixture model to the distribution of radial velocity jitter for
all similar sources selected through Step 1. This mixture model includes a normal
distribution to represent the intrinsic jitter for single stars, and a log-normal
distribution to represent stellar multiples. Specifically the model parameters include
a mixing weight parameter $\theta$ that is constrained between $(0, 1)$, the mean 
$\mu_{s}$ and standard deviation $\sigma_{s}$ of the normal distribution,
and the mean $\mu_{m}$ and standard deviation $\sigma_{m}$ of the 
log-normal distribution.

We assume a beta prior on $\theta$  such that $\theta \sim \mathcal{B}\left(5, 5\right)$
and require that $\sigma_{s}$ and $\sigma_{m}$ are larger than zero.
Depending on the distribution of the radial velocity jitter, we often found this mixture
challenging to optimize because the log-normal distribution could easily account for 
most of the observed distribution, which tended the mixing weight to zero and implied
that all selected sources were stellar multiples. To avoid this situation we placed a
constraint on the mean of the log-normal distribution to ensure that the \emph{mode}
of the log-normal distribution is larger than $\mu_{s} + 1\sigma_{s}$ such that
\begin{eqnarray}
%	\mu_{s} + 5\sigma_{s} & > & \exp\left(\mu_{m} - \sigma_{m}^2\right) \label{eq:rvm_lower_bound} \\
	\mu_{s} + 1\sigma_{s} & < & \exp\left(\mu_{m} - \sigma_{m}^2\right) \label{eq:rvm_upper_bound} \quad .
\end{eqnarray}

@@

Wherever possible we initialised the model parameters from nearby points (within the
ball selected by the k-d tree) where an optimised result already exists. This was primarily to minimise
computational cost; the initialisation did not have an impact on the result. 
We required the log probability and the gradient of the log
probability to reach machine precision before considering the optimisation
converged. After convergence, we queried whether any points within the ball of
similar stars had not been optimized. If no result existed, we would next move to that
point. We refer to this process as \emph{swarm optimisation}, which we conducted in 
parallel, and illustrate in Figure \ref{fig:swarm_optimisation}.

@@

\subsubsection*{Step 3: Evaluating probability and system properties}

\todo{Bayes factors for each dimension?}

\todo{source properties: semi-amplitude (easy)}

\todo{Can we recover period and eccentricity using The Joker?}

@@

\section{Radial velocity completeness as a function of \Gaia\ source properties}
\todo{Figures}

>>>>

Stellar multiplicity can be inferred from \Gaia\ data using radial velocity
measurements, astrometry, and photometry. These data are sensitive in different 
ways. Here we describe the data that we use from the second \Gaia\ data 
release \citep{Lindegren:2018} before we define our methods.

@@

For a single star, or a star without a significant mass companion, the source radial
velocity jitter represents the minimum uncertainty with which \Gaia\ can measure 
radial velocity for a star of that colour, apparent magnitude, and absolute magnitude.
Stars with bluer colours have fewer absorption lines with deeper wings, and fainter stars
have on average lower signal-to-noise (S/N) ratios, both of which result in noisier radial
velocity measurements. The absolute magnitude is similarly important, as giant stars have narrow absorption lines than main-sequence stars of the same temperature and colour. 
For sources that are stellar multiples, the radial velocity jitter will be the quadrature sum
of the expected jitter for the primary (if it were a single star), and a contribution from the
radial velocity semi-amplitude of the orbit.

Some sources in \Gaia\ DR2 do not have reported radial velocities even though they are
apparently bright enough and within a suitable \bp\ - \rp\ colour range. The two most
likely explanations are that either the \Gaia\ team identified the source to be a
double-lined spectroscopic binary, or the radial velocity error was larger than 20\,km\,s$^{-1}$,
and so the source radial velocity was removed. While it may be appealing to use the \emph{lack}
of radial velocity measurement as an indicator of stellar multiplicity, we show in Appendix A
that such an approach is problematic due to completeness, and other complications arising from
the initial \Gaia\ source catalog. However, in the following sections we show that for 
sources that do not have reported radial velocities -- and it has an apparent magnitude 
and colour that suggests it ought to have a radial velocity reported -- the astrometric 
jitter is often sufficient to identify stellar multiples.

@@

Many higher-order star systems are unresolved point sources in \Gaia, but the
stellar companion(s) may have sufficient mass to measurably affect the tangential 
velocity of the primary and produce a detectable jitter in the astrometric 
position. 
%There are many other sources of astrometric jitter, but here we show 
%that the principle effect of astrometric excess noise is due to stellar 
%companions.
We make use of the astrometric unit weight error $u$ 
\begin{eqnarray}
	u = \sqrt{\frac{\chi^2_{al}}{N_{obs,al} - 5}} \label{eq:astrometric_unit_weight_error}
\end{eqnarray}
\noindent{}where $\chi^2_{al}$ is the astrometric $\chi^2$ value
in the along-scan direction (column \texttt{astrometric\_chi2\_al}) and 
$N_{obs,al}$ is the number of astrometric observations in the along-scan
direction (column \texttt{astrometric\_n\_obs\_al}). 
The astrometric unit weight error is unaffected by the `degree of freedom bug'
that occurred during the processing of the the second \Gaia\ data release
\citep[e.g., see Appendix A1 of ][]{Lindegren:2018}, which resulted in 80\% of
sources having zero astrometric excess noise. 
The astrometric unit weight error can be interpreted similarly to how the
astrometric excess noise would be interpreted (e.g., large values are less
consistent with a single star model), but using the astrometric unit weight
error allows us to distinguish astrometric binaries in a much larger sample.
And like the radial velocity jitter, the unit weight astrometric error of a
single source must be considered in context with stars of similar colour 
and apparent magnitude (but not necessarily absolute magnitude). For lack of
a more descriptive term, throughout this work we will refer to the astrometric
unit weight error as the \emph{astrometric jitter}, similar to how we refer
to the standard deviation of radial velocities as the \emph{radial velocity jitter}.

@@

We restrict our work to the brightest $\sim{}10^7$ sources in \Gaia\ DR2.
About 70\% of sources in this subset have reported radial velocities, and
this subset is relatively close to the Sun. Specifically, we executed the
following Astronomical Data Query Language \citep[\texttt{ADQL};][]{someone} 
query on \ESA\ \Gaia\ archive \citep{esa_gaia_archive}:

@@

We assume that for a sample of sources with similar properties:
\begin{itemize}
	\item \bp\ - \rp\ colours;
	\item apparent \rp\ magnitude; and
	\item absolute \rp\ magnitude,
\end{itemize}
\noindent{}the distribution of jitter (in astrometry or radial velocity)
is a mixture of two components:

\begin{enumerate}
\item \emph{single-star systems}, where the jitter represents the minimum noise that 
	   \Gaia\ can measure for a source with those properties,

@@

We further assume that the jitter for a single star will change depending on the
source properties. For example, the mean radial velocity jitter of single stars 
with $\teff \approx 6000\,\textrm{K}$ will be higher than the jitter of single stars with
$\teff \approx 5000\,\textrm{K}$, all else being equal. For these reasons, in order to evaluate whether 
a source is more likely to be a single star or a stellar multiple, we must consider 
the jitter in context with other sources that have a similar properties (\bp\ - \rp\ colour,
apparent \rp\ magnitude, and absolute \rp\ magnitude).\footnote{The magnitude in any band would be sufficient. Throughout
this work we chose the \rp\ band because its wavelength range most closely overlaps 
with that of the Radial Velocity Spectrograph on board \Gaia.} However, we cannot reliably describe a quantitative
relationship for \emph{how} exactly the jitter of a single star is
expected to vary given some source properties. For these reasons
we chose to adopt a flexible non-parametric model for stellar multiplicity, primarily
conditioned on the assumption that if we select enough sources with \emph{similar}
properties then within that subset of similar stars there exists \emph{some} single stars 
and \emph{some} stellar multiples -- even if we do not necessarily know \emph{a priori}
which sources are single stars, and which are stellar multiples.

We adopt a two-step procedure for detecting stellar multiplicity. In the first step, for a
given \Gaia\ source of interest, we consider the jitter in context of similar sources  in 
order to measure the jitter of single stars and evaluate whether this source is more consistent with being a single star or stellar multiple. While we can repeat this process 
for every source, this procedure is largely independent between two sources: there is no
constraint requiring the estimated jitter to be smooth across source parameter space. At the second step we use a Gaussian process to model the jitter for single stars across
source parameter space, and enforce smoothness. We illustrate our procedure in Figure \ref{fig:schematic} and describe it in more detail below.

\subsection{Non-parametric model}

When considering the (radial velocity or astrometric) jitter  for a \Gaia\ source of interest,
we select $N$ sources in a `\emph{ball}' that have similar \bp\ - \rp\ colour, apparent \rp\ magnitude, and absolute \rp\ magnitudes as the source of interest. We chose $N = 1024$ in order to ensure that the sample is large enough such that
it will include \emph{some} single stars and some stellar multiples, while remaining small enough
such that all sources have very similar colours and magnitudes. There are other subtle constraints that we enforce on the sources in the `\emph{ball}'. Specifically we scale the relative dimensions such that 0.05\,mags in \bp\ - \rp\
colour is equivalent to 0.5\,mags in absolute \rp\ magnitude, and 0.25\,mags in apparent magnitude. We further require that the ball has a minimum radius of 0.1\,mags in \bp\ - \rp\ colour and 0.5\,mags in apparent and absolute magnitude. In dense regions of parameter space there are often more than $N$ sources that are within a multi-dimensional ball of that radius. In those situations we randomly chose $N$ sources within the ball such that the randomly chosen sources approximately span the minimum radius constraint.

Given that sample of $N$ similar sources, we fit a two-component mixture model to the
distribution of jitter values. We treat the astrometric and radial velocity jitter separately.  This mixture model includes a normal (Gaussian) distribution to represent the intrinsic
jitter for single stars, and a log-normal distribution to represent stellar multiples.
The model parameters include the relative mixing weight $\theta$, the mean $\mu_\textrm{s}$
and standard deviation $\sigma_\textrm{s}$ of the normal distribution, and the mean $\mu_\textrm{m}$ and standard deviation $\sigma_\textrm{m}$ of the log-normal distribution. We assume the priors
\begin{eqnarray}
\theta 				& \sim & \textrm{Uniform}(0, 1)    \\
\mu_\textrm{s} 		& \sim & \textrm{Uniform}(0.5, 10) \\
\sigma_\textrm{s} 	& \sim & \textrm{Uniform}(0, 10)   \\
\mu_\textrm{m} 		& \sim & \textrm{Uniform}(\log[\mu_\textrm{s} + \sigma_\textrm{s}] + \sigma_\textrm{m}^2, \infty) \label{eq:mode_prior} \\
\sigma_\textrm{m} 	& \sim & \textrm{Uniform}(0, 10)   \quad .
\end{eqnarray}

Where Eq. \ref{eq:mode_prior} ensures that the \emph{mode} of the log-normal
distribution must be greater than $\mu_\textrm{s} + 1\sigma_\textrm{s}$, thereby
requiring that stellar multiples must have higher jitter than single stars.

%We illustrate our approach in Figure \ref{fig:npm_schematic}, which can be qualitatively
%described by the following steps. When considering the radial velocity jitter for a
%\Gaia\ source of interest, we select $N$ sources that have a similar \bp\ - \rp\ colour,
%apparent \rp\ magnitude, and absolute \rp\ magnitude. We then fit the distribution of
%radial velocity jitter $\RVJitter$ as a mixture of two components: a normal distribution
%to represent single stars, and a log-normal distribution to represent stellar multiples.
%We perform these steps for $\approx10^4$ sources. While this process approximately maps
%how the radial velocity jitter of single (and multiple) stars varies across the Hertszprung-Russell
%diagram, each source of interest is considered independent from all others. For example,
%for three sources that are all indistinguishable from each other in \bp\ - \rp\ colour, 
%apparent \rp\ magnitude, and absolute \rp\ magnitude, the procedure we described above
%treats the fitting of these three mixture models as entirely independent of each other,
%even though we suspect that the parameters of the three mixture models ought to be 
%identical. In other words: the procedure as described does not enforce any smoothness.
%Given enough data this is not a problem for densely-populated regions of the 
%Hertszprung-Russell diagram as the optimized results \emph{will} be (statistically)
%indistinguishable for similar sources. In practice this poses a problem for poorly-sampled
%regions of the Hertzsprung-Russell diagram, and raises subjective questions on what
%constitutes \emph{the most} similar type of source.

@@

%\subsection*{Step 1: Selecting similar sources}

%We construct a $k$-dimensional tree \citep[k-d tree;][]{someone} for all sources in the
%dimensions of \bp\ - \rp\ colour, apparent \rp\ magnitude, and absolute \rp\ magnitude.
%When doing so we scale the dimensions such that \todo{0.1}\,mags in \bp\ - \rp\ colour is
%approximately equal to \todo{1} magnitude in apparent \rp\ magnitude and \todo{1} magnitude in 
%absolute \rp\ magnitude. We use this $k$-d tree for selecting a `ball' of similar
%stars for a given source.

%We enforce a number of constraints when selecting similar stars. For each source
%we require the multi-dimensional `ball' to include at least \todo{128} sources.
%We further require that the radius of the ball extends to at least \todo{0.05}\,mags in 
%\bp\ - \rp\ colour, \todo{0.5}\,mags in apparent \rp\ magnitude, and \todo{0.5}\,mags in absolute
%\rp\ magnitude. If the initial query of \todo{128} points does not extend wide enough to
%meet our constraints on minimum radius, we extend the radius (linearly in each scaled)
%dimension until the minimum constraints on radii are met. Our minimum radii constraints
%imply that there will be many (e.g., $>>\todo{128}$) points returned in dense regions of the
%Hertszprung-Russell diagram. In these situations we then subsample the selected points
%to return a random \todo{1024} sources in any ball, while enforcing our constraints on
%minimum radii. As a result, the number of `similar' sources selected will range from
%a minimum of \todo{128} to a maximum of \todo{1024}, and the minimum radii constraints on various
%dimensions are always maintained.

@@

%\subsection*{Step 2: Fitting the two-component mixture}

@@

\begin{comment}
Wherever possible we initialised the model parameters from nearby points (within the
ball selected by the k-d tree) where an optimised result already exists. This was primarily to minimise
computational cost; the initialisation did not have an impact on the result. 
We required the log probability and the gradient of the log
probability to reach machine precision before considering the optimisation
converged. After convergence, we queried whether any points within the ball of
similar stars had not been optimized. If no result existed, we would next move to that
point. We refer to this process as \emph{swarm optimisation}, which we conducted in 
parallel, and illustrate in Figure \ref{fig:swarm_optimisation}.

@@

\subsubsection*{Step 3: Evaluating probability and system properties}
\end{comment}

We performed this procedure and optimized this mixture model for $K = 5000$ \todo{randomly}
chosen sources, giving us noisy estimates of the astrometric and radial velocity jitter
for sources across the Hertszprung-Russell diagram. In theory we could have repeated the
procedure above for all sources, but we were computationally limited  by the number of
sources that we could use in a Gaussian process.

\subsection{Gaussian processes}

@@

\section{Radial velocity completeness as a function of \Gaia\ source properties}
